Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems

### 2025-02-19 paper/manuscript.Rmd
Incorporate Willem comments

Biodiversity changes are expected to profoundly impact ecosystem functioning and
stability, yet empirical evidence remains limited. Bridging this gap requires
aligning biodiversity trends with food web theory and Biodiversity-Ecosystem
Functioning research, which provide strong predictions about the ecological
consequences of biodiversity loss. Most insights rely on spatial comparisons of
ecological communities, assuming they reflect temporal trends—a space-for-time
substitution. Here, we analyze biodiversity and food web time series from over
400 riverine fish communities across France (1995–2018) to evaluate how
biodiversity trends reshape food web structure. We examine relationships among
species richness, community biomass, and three key food web metrics:
connectance, weighted average trophic level, and trophic pathway redundancy. We
also test the space-for-time hypothesis using spatial gradients and a
theoretical food web model. Structural equation models and robustness analyses
reveal that declines in species richness strongly correlate with reduced
community biomass, with top trophic-level species playing a disproportionate
role in food web changes. Declining biomass leads to the loss of top trophic
levels and decreased connectance, while species richness negatively correlates
with connectance. Both biomass and species richness declines reduce trophic
pathway redundancy, suggesting biodiversity loss undermines community robustness
to future perturbations. Our findings align with food web theory and
Biodiversity-Ecosystem Functioning research, support the validity of
space-for-time approaches for basic food web metrics, and highlight the utility
of food web metrics as indicators of ecosystem function and fragility.

@@

Human pressures are driving profound biodiversity changes, including freshwater
and terrestrial biodiversity declines linked to land degradation
[@newbold_global_2015; @cardinale_is_2018; @danet_past_2024] and reductions in
marine fish biomass due to intensified fishing pressures on predatory species
[@christensen_century_2014]. A central goal of Biodiversity-Ecosystem
Functioning and food web research has been to understand the ecological
consequences of these widespread biodiversity losses for ecosystem functioning
and stability [@loreau_biodiversity_2001; @loreau_ecological_2022;
@thebault_food-web_2003; @schneider_animal_2016; @dunne_network_2002;
@sole_complexity_2001; @allesina_functional_2009; @montoya_press_2009; @sanders_trophic_2018]. However,
research on biodiversity changes remains largely disconnected from studies
examining their ecological impacts, primarily due to a lack of simultaneous
temporal data on biodiversity and food web dynamics [@danet_species_2021]. As a
result, much of our understanding of the ecological consequences of biodiversity
changes over time relies on experimental and theoretical approaches. These
approaches often depend on the critical assumption that spatial comparisons of
ecological communities can serve as a proxy for temporal trends—a space-for-time
substitution—when direct temporal data are scarce. To advance our understanding
of the ecological consequences of biodiversity changes, it is essential to
investigate how these changes are reshaping food web structure in natural
settings.

@@

Our food web reconstruction delivered `r nb_op` annual estimates of species
richness, biomass, connectance, weighted average trophic level, and the
proportion of redundant trophic links (hereafter “redundancy”, see Appendix B)
from our `r nb_st` fish communities with 10+ years of population dynamics
spanning 1995-2018 over France (Fig. \@ref(fig:matmet)a-c). We implemented Bayesian hierarchical
models to jointly estimate the temporal trends in these metrics. We found a
great heterogeneity in the temporal trends at the site level (community biomass:
from `r grbmn`% to +`r grbmx`% per decade, species richness: `r grrhn` to
+`r grrhx`%, average trophic level: `r grtln` to +`r grtlx`, connectance:
`r grctn` to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig. S1) which was
effectively captured by the hierarchical model ( conditional R2 ranging from
`r r2cond$log_bm_std` for community biomass to `r r2cond$log_rich_std` for
species richness, Table S2).

@@

The total effect of the temporal trends in species richness on the temporal
trend in connectance was negative ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a), resulting from a negative direct effect ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f) and a positive indirect effect via changes
in the temporal trends in community biomass ($r_\delta =$ `r ridtps$ct$r`).
Thus, when communities experienced an increase, or decrease, in species
richness, they also experienced a concomitant increase, or decrease, in
connectance. In contrast, species richness temporal trends had no direct effects
on trends in average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig.
\@ref(fig:sem)g), but a positive indirect effect via changes in the trends of
community biomass ($r_\delta =$ `r ridtps$tlvl$r`), which
adds up to a total positive effect of species richness temporal trends on
average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig. \@ref(fig:std)a).
Temporal trends in species richness had a strong total positive effect on the
trends in redundancy (resp. $r_\delta =$ `r rdtps$red$r`, `r ridtps$red$r`), via
direct and indirect positive effects ($r_\delta =$ `r rttps$red$r`). All
together, it means that communities that have experienced species richness
declines over the last decades have decreased in average trophic level and
redundancy but increased in connectance. Conversely, those that experienced
species richness increases have increased in average trophic level, redundancy,
but decreased in connectance.

The effects of species richness were more complex, because species richness in
this model has indirect effects on food web structure through community biomass.
Species richness had a direct negative effect on connectance ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f), a positive indirect effect through
community biomass ($r_\delta =$ `r ridtps$ct$r`), so that species richness had a
negative effect in total on connectance ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a). In contrast, species richness had no direct effects on in
average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig. \@ref(fig:sem)g), a
positive indirect effect through community biomass ($r_\delta =$
`r ridtps$tlvl$r`), which adds up to a positive total effect of species richness
temporal trends on average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig.
\@ref(fig:std)a). Species richness temporal trends had direct and indirect
positive effect on the temporal trends in redundancy (resp. $r_\delta =$
`r rdtps$red$r`, `r ridtps$red$r`), which added up to a strong positive effect
($r_\delta =$ `r rttps$red$r`).

@@

We found strong agreement between our temporal and spatial analyses, as well as
with predictions from food web theory and Biodiversity-Ecosystem Functioning
research. This suggests that spatial comparisons of generic topological food web
properties and biodiversity metrics may serve as reasonable surrogates for
predicting temporal trends—supporting the validity of the space-for-time
substitution in both empirical data and bioenergetic food web models. However,
this agreement did not hold for the relationship between weighted average
trophic level (hereafter “average trophic level”) and species richness. This
discrepancy may arise because average trophic level incorporates both food web
topology and biomass distribution, which are influenced by dynamic processes
such as bottom-up and top-down control. These processes likely require
additional information about species metabolism and productivity
[@brose_allometric_2006; @barbier_pyramids_2019]. This interpretation is
supported by the fact that our spatial analysis accurately predicted the
temporal association between community biomass and average trophic level.

@@

We first estimated the body mass of each fish individual with the allometric
law: $B_i = 0.01 \times l_i^{3.03}$, $B_i$ being the biomass expressed in gram and $l_i$ the
body length of the individual $i$ expressed in millimeters [@froese_cube_2006;
@danet_species_2021]. Then the biomass of each species and community
were obtained by summing the body mass of individuals. Finally, the biomass was
reported to the sampled surface (in $m^2$) to account for sampling effort.

@@

We inferred the metaweb according to ontogenic diet data and body size
[@gravel_inferring_2013; @poisot_trophic_2013; @brose_predator_2019;
@bonnaffe_comparison_2021; @danet_species_2021]. The method is extensively
described in @bonnaffe_comparison_2021 and @danet_species_2021. The food diet of
stream fishes changes along their life cycle or stage, therefore according to
their body size. Each fish species was divided in nine size classes, i.e.
trophic species, equally distributed in their observed body size range. Thus,
a trophic species is a group of individuals of the same species and size class
that are expected to share the same preys and predators because trophic
interactions are largely determined by predator-prey size ratio
[@brose_allometric_2006; @brose_predator_2019]. We also added seven nodes present in the ontogenic
food diet database: detritus, biofilm, phytoplankton, zooplankton, macrophages,
phytobenthos and zoobenthos.  The resulting metaweb contained 45 fish species
divided into 9 size classes plus 7 resource nodes, summing up to 412 nodes.
Consideration of size classes, i.e.  trophic species instead of species,
improves inference of trophic interactions [@jennings_weak_2001].

@@

At each site, yearly food webs were described by their community biomass and
species richness. Yearly food web structure was described by their connectance,
describing the degree of generalism in a network [@dunne_network_2006;
@delmas_analysing_2019]. Connectance was computed as the ratio between the
number of actual links and the number of potential links ($C =
\frac{L}{S(S-1)}$, with $L$ being the number of links and $S$ the number of
trophic species). We also computed the average trophic level, weighted by the
biomass of the trophic species. The trophic level of a trophic species was
computed recursively as 1 added to the mean trophic level of the preys it
consumes [@herendeen_ecological_2008]. We computed the number of redundant and
functional links using the dominator tree approach [@allesina_functional_2009].
A species $i$ is a dominator of species $j$ if $i$ mediates the connection $j$
to basal resources such as the extinction of $i$ remove the connection of $j$ to
basal resources. It is then to identify the set of trophic links that are
essential to connect species to basal resources, i.e. functional links, and the
redundant links. This method requires a single basal root in the food webs. To
do so, we added a basal dummy resource node to the empirical food webs connected
to biofilm and phytoplankton resource nodes. For food webs generated from
theoretical simulations (see below), we added a resource node connected to all
primary producers. We computed the set of dominator trees using the
`dominator_tree()` function from the `igraph` R package (Appendix B). We
computed the number of functional links as the number of links in the resulting
dominator tree of the food web. We checked that the procedure reproduced the
examples presented in previous studies [@allesina_functional_2009, Appendix B].
We then derived the proportion of redundant links in food webs of empirical and
theoretical simulations.

@@

We modeled the temporal trends of food web structure at each site ($Y_i$) using
linear hierarchical model. Time was considered as the number of years since the
first sampling ($t = 0$) at each site. We accounted for the spatial structure of
the data by adding random intercepts ($a$) and slopes ($b$) for the
basin ($a_n$, $b_n$) and site ($a_{i|n}$, $b_{i|n}$) identity, site effects
being nested in the basin one.

@@

The model parameters were estimated using INLA (Integrated Nested Laplacian
Approximation) Bayesian inference with default uninformative priors
[@rue_bayesian_2017]. The prior distributions for the fixed intercept and slope were
modeled as a flat normal distribution with a mean of zero ($\mathcal{N}(0,\, 1000)$).
Prior distributions for the Gaussian error $\epsilon_{i|n, t}$ and the random
effects were modeled as a log gamma distribution with
shape and inverse scale parameters ($\mathcal{G}(s, \, \tau) = \mathcal{G}(1, \,
5.10^-5)$), then their respective standard deviation were obtained by
back-transforming the inverse scale parameter ($\sigma = 1 / \sqrt{\tau}$).
Community biomass and species richness were log-transformed.

The performance of the model was assessed visually by plotting the fitted versus
the observed values (Fig. S5). We also visually inspected the Posterior
Integral Transform (PIT) and Conditional Predictive Ordinate (CPO) to assess the
quality of the model fit. We also computed the marginal ($R^2_m$) and conditional
($R^2_c$) R-squared of the models following previous formulations
[@nakagawa_general_2013; @lahuis_explained_2014]. In order to account for the
variability associated with the priors [@gelman_r-squared_2019], we computed the
variance of the posterior distribution  of each predicted values ($Var_{fit} =
\sigma^2(\hat{Y_i})$). We computed $R^2$ values for each predicted values, then
obtaining a $R^2$ distribution [@gelman_r-squared_2019]. We reported mean
marginal and conditional $R^2$ associated the 95% credible interval computed
using the Highest Posterior Density method.

@@

We extracted the temporal trends at each site from the hierarchical model
using mean BLUP values (Best Linear Unbiaised Prediction), then multiplied
by 10 to get temporal trends by decade. For community biomass and species
richness which were log-transformed, we derived temporal trends in percentage
change by decade by back-transforming the slopes ($\beta\prime = (e^{\beta } - 1)
\times 100 \times 10$).

@@

In order to control for the uncertainties of the estimated temporal trends, we
weighted each temporal trend by the inverse of the sum of the standard deviation
of the estimated trends, i.e. the estimated slopes for the response and the
explicative variables ($w_j = 1 / \sum{e_j}$, $w_j$ being the weight, and
$s_{e_j}$ the standard deviation on the temporal trends of the variables).

The coefficients of the Structural Equation Models were computed with
PiecewiseSEM R package [@lefcheck_piecewisesem_2016]. The coefficients were
standardized and the confidence intervals were computed with semEff R package
[@murphy_semeff_2021]. Confidence intervals were obtained by bootstrapping the
linear models 1000 times and computing adjusted bootstrap percentile intervals
(BCa), which are adjusted for the asymmetric distribution of resampling. The marginal
and conditional $R^2$ associated to the linear models of the SEMs were computed
with the Nagelkerke method [@lefcheck_piecewisesem_2016, Table S4].

@@

We simulated extinction sequences in the metaweb according to three scenarios to
get expectations on how the food web structure should change according to the
addition and deletion of species and nodes. The scenario of extinction sequences
were related to the trophic level of the species/nodes. In the decreasing and
increasing scenario, we removed species/nodes in decreasing, resp. increasing
order, of their trophic level, i.e. respectively beginning by the highest and
lowest trophic level. In the random extinction sequence, we built 50 random
extinction sequences. For each simulation, we removed fish species/nodes that
lost all their preys. We then computed the connectance, the average and the
maximum trophic level after each extinction.

@@

We used a bioenergetic food web model to simulate the effects of community
biomass and species richness changes on food web structure. This food web model
lifts one of the limitations of our data, that is it includes the biomass of all
the food webs from primary producers to top predators. This model allows
simulating complex communities because its parameterization depends mainly on
species bodymasses. We first generated `r nfw` food webs varying in species
richness (range: `r s_range["min"]`-`r s_range["max"]`)
and connectance (range: `r c_range["min"]`-`r c_range["max"]`) using the niche
model algorithm, which reproduces food web topology typically found in empirical
food webs [@williams_simple_2000]. We set species bodymasses ($M_i$) using a predator-prey
mass ratio of 100 ($Z$) [@brose_allometric_2006; @brose_predator_2019], such as $M_i =
Z^{(TL_i - 1)}$, $TL_i$ being the trophic level of the species $i$. The
metabolic rates of the species ($x_i$) were then derived from allometric
relationships between metabolism and bodymass empirically estimated
[@brose_allometric_2006], such as $x_i = -0.314M_i^{-0.25}$.

@@

The primary producer biomass increase with a growth rate ($r_i$) of 1 and a
logistic growth (eq. \@ref(eq:growth)), growth being limited by intraspecific
($\alpha_{ii}$, set to 1), interspecific ($\alpha_{ij}$, set to 0.5) competition,
and carrying capacity ($K_i$, varying in the simulations). Primary producers
loses biomass by being eaten by consumers at a rate that depend on consumer
metabolic rates ($x_j$), attack rates ($y_j$, set to 8 @delmas_analysing_2019),
a functional response ($F_{ji}$), and assimilation efficiency ($e_{ij}$, set to
0.45 for herbivores and 0.85 for carnivores, @delmas_analysing_2019).

@@

Consumers gain biomass by eating other species and lose biomass by being
consumed and by metabolic loses ($x_i$, eq. \@ref(eq:cons)). The functional
response (eq. \@ref(eq:holling)) of consumer feeding depends on the preference
for its preys ($\omega_{ij}$, set even across consumer preys) and
half-saturation rate ($B_0$, set to 0.5). The hill coefficient ($h$, set to 1.5)
determines the shape of the functional response (Holling's type II or III).

>>>>

Biodiversity temporal changes are expected to profoundly impact ecosystem
functioning and stability, yet large-scale empirical evidence remains limited. Bridging this
gap requires aligning biodiversity temporal trends with food web theory and
Biodiversity-Ecosystem Functioning research, which provide strong predictions
about the ecological consequences of biodiversity loss. Most insights rely on
spatial comparisons of ecological communities, assuming they reflect temporal
trends—a space-for-time substitution. Here, we analyze biodiversity and food web
time series from over 400 riverine fish communities across France (1995–2018) to
evaluate how biodiversity temporal trends reshape food web structure. We examine
relationships among species richness, community biomass, and three key food web
metrics: connectance, weighted average trophic level, and trophic pathway
redundancy. We also test the space-for-time hypothesis using spatial gradients
and a theoretical food web model. Structural equation models and robustness
analyses reveal that declines in species richness strongly correlate with
reduced community biomass, with top trophic-level species playing a
disproportionate role in food web changes. Declining biomass is associated with the loss
of top trophic levels and decreased connectance, while species richness
negatively correlates with connectance. Both biomass and species richness
declines reduce trophic pathway redundancy, suggesting biodiversity loss
undermines community robustness to future perturbations. Our findings align with
food web theory and Biodiversity-Ecosystem Functioning research, support the
validity of space-for-time approaches for basic food web metrics, and highlight
the utility of food web metrics as indicators of ecosystem function and
fragility.

@@

Human pressures are driving profound biodiversity changes, including freshwater
and terrestrial biodiversity declines linked to land degradation
[@newbold_global_2015; @cardinale_is_2018; @danet_past_2024] and reductions in
marine fish biomass due to intensified fishing pressures on predatory species
[@christensen_century_2014]. A central goal of Biodiversity-Ecosystem
Functioning and food web research has been to understand the ecological
consequences of these widespread biodiversity losses for ecosystem functioning
and stability [@loreau_biodiversity_2001; @loreau_ecological_2022;
@thebault_food-web_2003; @schneider_animal_2016; @dunne_network_2002;
@sole_complexity_2001; @allesina_functional_2009; @montoya_press_2009; @sanders_trophic_2018]. However,
research on biodiversity changes remains largely disconnected from studies
examining their ecological impacts, primarily due to a lack of simultaneous
temporal data on biodiversity and food web dynamics [@danet_species_2021]. As a
result, much of our understanding of the ecological consequences of biodiversity
changes over time relies on experimental and theoretical approaches. These
approaches often rely on the critical assumption that spatial comparisons of
ecological communities can serve as a proxy for temporal trends—a space-for-time
substitution—when direct temporal data are scarce. To advance our understanding
of the ecological consequences of biodiversity changes, it is essential to
investigate how these changes are reshaping food web structure in natural
settings.

@@

Our food web reconstruction delivered `r nb_op` annual estimates of species
richness, biomass, connectance, weighted average trophic level, and the
proportion of redundant trophic links (hereafter “redundancy”, see Appendix B)
from our `r nb_st` fish communities with 10+ years of population dynamics
spanning 1995-2018 over France (Fig. \@ref(fig:matmet)a-c). We implemented
Bayesian hierarchical models accounting for spatial variation at both
hydrographic river basin and site levels to jointly estimate the temporal trends in these metrics. We found a
great heterogeneity in the temporal trends at the site level
(community biomass: from `r grbmn`% to +`r grbmx`% per decade, species richness:
`r grrhn` to +`r grrhx`%, average trophic level: `r grtln` to +`r grtlx`,
connectance: `r grctn` to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig.
S1) which was effectively captured by the hierarchical model ( conditional R2
ranging from `r r2cond$log_bm_std` for community biomass to `r
r2cond$log_rich_std` for species richness, Table S2).

@@

The total effect of the temporal trends in species richness on the temporal
trend in connectance was negative ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a), resulting from a negative direct effect ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f) and a positive indirect effect via changes
in the temporal trends in community biomass ($r_\delta =$ `r ridtps$ct$r`).
Thus, when communities experienced an increase, or decrease, in species
richness, they also experienced a concomitant decrease, or increase, in
connectance. In contrast, species richness temporal trends had no direct effects
on trends in average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig.
\@ref(fig:sem)g), but a positive indirect effect via changes in the trends of
community biomass ($r_\delta =$ `r ridtps$tlvl$r`), which
adds up to a total positive effect of species richness temporal trends on
average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig. \@ref(fig:std)a).
Temporal trends in species richness had a strong total positive effect on the
trends in redundancy (resp. $r_\delta =$ `r rdtps$red$r`, `r ridtps$red$r`), via
direct and indirect positive effects ($r_\delta =$ `r rttps$red$r`). All
together, this means that communities that have experienced species richness
declines over the last decades have decreased in average trophic level and
redundancy but increased in connectance. Conversely, those that experienced
species richness increases have increased in average trophic level, redundancy,
but decreased in connectance.

The effects of species richness were more complex, because species richness in
this model has indirect effects on food web structure through community biomass.
Species richness had a direct negative effect on connectance ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f), a positive indirect effect through
community biomass ($r_\delta =$ `r ridtps$ct$r`), so that species richness had a
negative effect in total on connectance ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a). In contrast, species richness had no direct effects on
average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig. \@ref(fig:sem)g), a
positive indirect effect through community biomass ($r_\delta =$
`r ridtps$tlvl$r`), which adds up to a positive total effect of species richness
temporal trends on average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig.
\@ref(fig:std)a). Species richness temporal trends had direct and indirect
positive effect on the temporal trends in redundancy (resp. $r_\delta =$
`r rdtps$red$r`, `r ridtps$red$r`), which added up to a strong positive effect
($r_\delta =$ `r rttps$red$r`).

@@

We found strong agreement between our temporal and spatial analyses, as well as
with predictions from food web theory and Biodiversity-Ecosystem Functioning
research. This suggests that spatial comparisons of generic topological food web
properties and biodiversity metrics may serve as reasonable surrogates for
predicting temporal trends—supporting the validity of the space-for-time
substitution in both empirical data and bioenergetic food web models. However,
this agreement did not hold for the relationship between weighted average
trophic level (hereafter “average trophic level”) and species richness. This
discrepancy may arise because average trophic level incorporates both food web
topology and biomass distribution, which are influenced by dynamic processes
such as bottom-up and top-down control. These processes likely require
additional information about species metabolism and productivity
[@brose_allometric_2006; @barbier_pyramids_2019]. This interpretation is
supported by the fact that our spatial analysis accurately predicted the
temporal association between community biomass and average trophic level.
Importantly, this finding highlights that the relationship between
biodiversity—biomass and species richness—and food web
characteristics—connectance, trophic level, and redundancy—is not purely
structural but also functional. Indeed, predictions of food web structure are
most accurate when both structural (species richness) and functional (biomass)
metrics are considered.

@@

We first estimated the body mass of each fish individual with the allometric
law: $B_i = 0.01 \times l_i^{3.03}$, $B_i$ being the biomass expressed in gram and $l_i$ the
body length of the individual $i$ expressed in millimeters [@froese_cube_2006;
@danet_species_2021]. Then the biomass of each species and community
were obtained by summing the body mass of individuals. This resulted in a
biomass score for the whole community of each site. Finally, the biomass was
reported to the sampled surface (in $m^2$) to account for sampling effort.

@@

We inferred the metaweb according to ontogenic diet data and body size
[@gravel_inferring_2013; @poisot_trophic_2013; @brose_predator_2019;
@bonnaffe_comparison_2021; @danet_species_2021]. The method is extensively
described in @bonnaffe_comparison_2021 and @danet_species_2021. The food diet of
stream fishes changes along their life cycle or stage, therefore according to
their body size. Each fish species was divided in nine size classes, i.e.
trophic species, equally distributed in their observed body size range. Thus,
a trophic species is a group of individuals of the same species and size class
that are expected to share the same prey and predators because trophic
interactions are largely determined by predator-prey size ratio
[@brose_allometric_2006; @brose_predator_2019]. We also added seven nodes present in the ontogenic
food diet database: detritus, biofilm, phytoplankton, zooplankton, macrophages,
phytobenthos and zoobenthos.  The resulting metaweb contained 45 fish species
divided into 9 size classes plus 7 resource nodes, summing up to 412 nodes.
Consideration of size classes, i.e.  trophic species instead of species,
improves inference of trophic interactions [@jennings_weak_2001].

@@

At each site, yearly food webs were described by their community biomass and
species richness. Yearly food web structure was described by their connectance,
describing the degree of generalism in a network [@dunne_network_2006;
@delmas_analysing_2019]. Connectance was computed as the ratio between the
number of actual links and the number of potential links ($C =
\frac{L}{S(S-1)}$, with $L$ being the number of links and $S$ the number of
trophic species). We also computed the average trophic level, weighted by the
biomass of the trophic species. The trophic level of a trophic species was
computed recursively as 1 added to the mean trophic level of the preys it
consumes [@herendeen_ecological_2008]. We computed the number of redundant and
functional links using the dominator tree approach [@allesina_functional_2009].
A species $i$ is a dominator of species $j$ if $i$ mediates the connection of $j$
to basal resources such that the extinction of $i$ removes the connection of $j$ to
basal resources. It is then used to identify the set of trophic links that are
essential to connect species to basal resources, i.e. functional links, and the
redundant links. This method requires a single basal root in the food webs. To
do so, we added a basal dummy resource node to the empirical food webs connected
to biofilm and phytoplankton resource nodes. For food webs generated from
theoretical simulations (see below), we added a resource node connected to all
primary producers. We computed the set of dominator trees using the
`dominator_tree()` function from the `igraph` R package (Appendix B). We
computed the number of functional links as the number of links in the resulting
dominator tree of the food web. We checked that the procedure reproduced the
examples presented in previous studies [@allesina_functional_2009, Appendix B].
We then derived the proportion of redundant links in food webs of empirical and
theoretical simulations.

@@

We modeled the temporal trends of food web structure at each site ($Y_i$) using
linear hierarchical model which accounts for the hierarchical spatial structure
of the data, i.e. the sites nested in hydrographic basins. Time was considered as the number of years since the
first sampling ($t = 0$) at each site. $\alpha$ is the intercept and $\beta$ the
slope estimating the temporal trends. The intercept and the slope parameters
have a fixed ($\alpha_0$, $\mu$) part and a random part. We accounted for the
spatial structure of the data by adding random intercepts ($a$) and slopes ($b$)
for the basin ($a_n$, $b_n$) and site ($a_{i|n}$, $b_{i|n}$) identity, site
effects being nested in the basin one.

@@

The model parameters were estimated using INLA (Integrated Nested Laplacian
Approximation) Bayesian inference with default uninformative priors
[@rue_bayesian_2017]. The prior distributions for the fixed intercept ($\alpha_0$) and slope ($\mu$) were
modeled as a flat normal distribution with a mean of zero ($\mathcal{N}(0,\, 1000)$).
Prior distributions for the Gaussian error $\epsilon_{i|n, t}$ and the random
effects were modeled as a log gamma distribution with
shape and inverse scale parameters ($\mathcal{G}(s, \, \tau) = \mathcal{G}(1, \,
5.10^-5)$), then their respective standard deviation were obtained by
back-transforming the inverse scale parameter ($\sigma = 1 / \sqrt{\tau}$).
Community biomass and species richness were log-transformed.

The performance of the model was assessed visually by plotting the fitted versus
the observed values (Fig. S5). We also computed the marginal ($R^2_m$) and conditional
($R^2_c$) R-squared of the models following previous formulations
[@nakagawa_general_2013; @lahuis_explained_2014]. In order to account for the
variability associated with the priors [@gelman_r-squared_2019], we computed the
variance of the posterior distribution  of each predicted values ($Var_{fit} =
\sigma^2(\hat{Y_i})$). We computed $R^2$ values for each predicted values, then
obtaining a $R^2$ distribution [@gelman_r-squared_2019]. We reported mean
marginal and conditional $R^2$ associated the 95% credible interval computed
using the Highest Posterior Density method.

@@

We extracted the temporal trends at each site from the hierarchical model
using mean BLUP values (Best Linear Unbiased Prediction), then multiplied
by 10 to get temporal trends by decade. For community biomass and species
richness which were log-transformed, we derived temporal trends in percentage
change by decade by back-transforming the slopes ($\beta\prime = (e^{\beta } - 1)
\times 100 \times 10$).

@@

In order to control for the uncertainties of the estimated temporal trends, we
weighted each temporal trend by the inverse of the sum of the standard deviation
of the estimated trends, i.e. the estimated slopes for the response and the
explicative variables ($w_j = 1 / \sum{s_j}$, $w_j$ being the weight, and
$s_j$ the standard deviation on the temporal trends of the variables).

The coefficients of the Structural Equation Models were computed with
PiecewiseSEM R package [@lefcheck_piecewisesem_2016]. The coefficients were
standardized and the confidence intervals were computed with semEff R package
[@murphy_semeff_2021]. Confidence intervals were obtained by bootstrapping the
linear models 1000 times and computing adjusted bootstrap percentile intervals
(BCa), which are adjusted for the asymmetric distribution of resampling. The marginal
and conditional $R^2$ associated with the linear models of the SEMs were computed
with the Nagelkerke method [@lefcheck_piecewisesem_2016, Table S4].

@@

We simulated extinction sequences in the metaweb according to three scenarios to
get expectations on how the food web structure should change according to the
addition and deletion of species and nodes. The scenario of extinction sequences
were related to the trophic level of the species/nodes. In the decreasing and
increasing scenario, we removed species/nodes in decreasing, resp. increasing
order, of their trophic level, i.e. respectively beginning by the highest and
lowest trophic level. In the random extinction sequence, we built 50 random
extinction sequences. For each simulation, we removed fish species/nodes that
lost all their prey and iterate this removal until all remaining species have at
least one prey. We then computed the connectance, the average and the maximum
trophic level after each extinction.

@@

We used a bioenergetic food web model to simulate the effects of community
biomass and species richness changes on food web structure. This food web model
lifts one of the limitations of our data, that is it includes the biomass of all
the food webs from primary producers to top predators. This model allows
simulating complex communities because its parameterization depends mainly on
species body masses. We first generated `r nfw` food webs varying in species
richness (range: `r s_range["min"]`-`r s_range["max"]`)
and connectance (range: `r c_range["min"]`-`r c_range["max"]`) using the niche
model algorithm, which reproduces food web topology typically found in empirical
food webs [@williams_simple_2000]. We set species body masses ($M_i$) using a predator-prey
mass ratio of 100 ($Z$) [@brose_allometric_2006; @brose_predator_2019], such as $M_i =
Z^{(TL_i - 1)}$, $TL_i$ being the trophic level of the species $i$. The
metabolic rates of the species ($x_i$) were then derived from allometric
relationships between metabolism and body mass empirically estimated
[@brose_allometric_2006], such as $x_i = 0.88M_i^{-0.25}$.

@@

The primary producer biomass increase with a growth rate ($r_i$) of 1 and a
logistic growth ($G_i$, eq. \@ref(eq:growth)), growth being limited by intraspecific
($\alpha_{ii}$, set to 1), interspecific ($\alpha_{ij}$, set to 0.5) competition,
and carrying capacity ($K_i$, varying in the simulations). Primary producers
lose biomass by being eaten by consumers at a rate that depend on consumer
metabolic rates ($x_j$), attack rates ($y_j$, set to 4
@lajaaiti_ecologicalnetworksdynamicsjl_nodate), a functional response
($F_{ji}$), and assimilation efficiency which accounts for the fact that
consumers assimilate better the primary producer biomass than the consumer ones
($e_{ij}$, set to 0.45 for herbivores and 0.85 for carnivores,
@lajaaiti_ecologicalnetworksdynamicsjl_nodate).

@@

Consumers gain biomass by eating other species and lose biomass by being
consumed and by metabolic losses ($x_i$, eq. \@ref(eq:cons)). The functional
response (eq. \@ref(eq:holling)) of consumer feeding depends on the preference
for its preys ($\omega_{ij}$, set even across consumer preys) and
half-saturation rate ($B_0$, set to 0.5). The hill coefficient ($h$, set to 1.5)
determines the shape of the functional response (Holling's type II or III).
